---
title: "PA1_template"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reproducible Research Course Project 1
=====================================
## Introdution
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](https://www.fitbit.com/global/us/home), [Nike Fuelband](https://www.nike.com/help/a/why-cant-i-sync), or [Jawbone](https://www.jawbone.com/up) Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.  

This project makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

### Setting global option to turn warnings off

```{r}
knitr::opts_chunk$set(warning=FALSE)
```

### load data and chacteristics

```{r}
library(ggplot2)

activity <- read.csv("./activity.csv")

activity$date <- as.POSIXct(activity$date, "%Y-%m-%d")
```

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as <span style="color:red">NA</span>)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of `r nrow(activity)` observations in this dataset.

```{r}
weekday <- weekdays(activity$date)
activity <- cbind(activity,weekday)

summary(activity)
```

## Assignment

This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a **single R markdown** document that can be processed by **knitr** and be transformed into an HTML file.

Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use <span style="color:green">echo = TRUE</span> so that someone else will be able to read the code. **This assignment will be evaluated via peer assessment so it is essential that your peer evaluators be able to review the code for your analysis.**

For the plotting aspects of this assignment, feel free to use any plotting system in R (i.e., base, lattice, ggplot2)

Questions to be answered:

 + What is mean total number of steps taken per day?
 + What is the average daily activity pattern?
 + Imputing missing values
 + Are there differences in activity patterns between weekdays and weekends?

### 1. number of steps taken per day
```{r}
total_steps <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
names(total_steps) <- c("date", "steps")
hist(total_steps$steps, main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", 
     col = "grey", ylim = c(0,20), breaks = seq(0,25000, by=2500))
options(digits = 3)
mean_steps <- mean(total_steps$steps)
med_steps <- median(total_steps$steps)
```

the **mean** of the total steps taken by day is `r mean_steps`and the **median** is `r med_steps`

### 2. the average daily pattern average

* Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
int_avg <- tapply(activity$steps, activity$interval, mean, na.rm=TRUE, simplify=T)
df_ia <- data.frame(interval=as.integer(names(int_avg)), avg=int_avg)
names(df_ia) <- c("interval", "mean")
plot(df_ia$interval, df_ia$mean, type = "l", col="darkblue", lwd = 2, xlab="Interval", ylab="Average number of steps", main="Average number of steps per intervals")
max_steps <- df_ia[which.max(df_ia$mean), ]$interval
```

In the 5-minute interval, on average across all the days in the dataset, the maximum number of steps is : `r max_steps`

### 3. Imputing missing values

* there are a number of days/intervals where there are missing values (coded as <span style = "color:purple">NA</span>. The presence of missing days may introduce bias into some calculations or summaries of the data.
```{r}
NA_steps <- sum(is.na(activity$steps))
```

* the total number of <span style = "color:purple">NA</span> of rows is `r NA_steps`

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* the strategy consists of filling the missing values with mean of the steps used in previous step.

```{r}
released_steps <- df_ia$mean[match(activity$interval, df_ia$interval)]
```

Now, we fill and create a new dataset without <span style = "color:red">NA</span> values.

```{r}
new_activity <- transform(activity, steps = ifelse(is.na(activity$steps), yes = released_steps, no = activity$steps))
total_steps_unreleased <- aggregate(steps ~ date, new_activity, sum)
names(total_steps_unreleased) <- c("date", "daily_steps")
```

we make a histogram of the total number of steps taken each day and we calculate and report the **mean** and **median** total number of steps taken per day.

```{r}
hist(total_steps_unreleased$daily_steps, col = 'grey', 
     xlab = 'Total steps per day', ylim = c(0,30), 
     main = 'Total number of steps taken per day', 
     breaks = seq(0,25000, by = 2500))
mean_tt_stepun <- mean(total_steps_unreleased$daily_steps)
med_tt_stepun <- median(total_steps_unreleased$daily_steps)
```

the new mean is `r mean_tt_stepun`  which more than the previous mean with <span style =  "color:red">NA</span> `r mean_steps` and even the median has became `r med_tt_stepun` from `r med_steps` with <span style =  "color:red">NA</span> values.

### 4. differences in activity patterns between weekdays and weekends

* we create a new variable in the dateset for weekends and weekdays. 

```{r}
new_activity$date <- as.Date(strptime(new_activity$date, format="%Y-%m-%d"))
library(dplyr)
activity1<-new_activity%>%
        mutate(dayType= ifelse(weekdays(new_activity$date)=="Saturday" | weekdays(new_activity$date)=="Sunday", "Weekend", "Weekday"))
head(activity1)
```

Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
average_stepact<-activity1 %>%
  group_by(dayType, interval) %>%
  summarize(av_stepday=sum(steps))
head(average_stepact)
library(lattice)
with(average_stepact, 
      xyplot(av_stepday ~ interval | dayType, 
      type = "l", 
      layout = c(1,2),
      main = "Total Number of Steps within Intervals by dayType",
      xlab = "Daily Intervals",
      ylab = "Average Number of Steps",
      lty = 1))
```